---
title: "Rmd exploration"
author: Miguel Prieto
date: Sept 2019
output: 
  ioslides_presentation:
    incremental: true
---

## Set options of knitting

- Just below the title of the Rmd document 

- output:\
  ioslides_presentation:\
    incremental:true


# Printing data

```{r, echo=FALSE, warning=FALSE, results = 'hide', message=FALSE}

## chunk options indicate if it must show code, warnings, results and messages

packages <- c("rmarkdown", "DT", "tibble", "gapminder")

lapply(packages, require, character.only = TRUE) # way to require multiple packages in one call

# browseVignettes(package = "rmarkdown")

```
## A tibble named gapminder

```{r, }
gapminder
```

--- 

```{r, include=FALSE, eval=FALSE}

--- # allows you to create a blank title slide
# symbol for section slide
  (##) for new slide with title

```

Convert mtcars dataframe to a tibble

```{r}

tibble::as_tibble(mtcars) 
```

---
1. datatable function shows a graphically attractive table
1. can be used on a tibble
```{r}
datatable(as_tibble(mtcars))

```


make a comment inline
<!--this is an inline comment, does not appear in output-->

## YAML modified

Settings go below the output document\
Change the teme with theme:____\
Add a table of contents with toc: true\
Make the toc float: toc_float: true.