---
title: "02102019-Data visualization(ggplot)_R for data science"
author: "MDP"
date: "10/2/2019"
output:  
  html_document:
    theme: flatly
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Package ggplot2 uses **"grammar of graphics"** to describe and build graphs

Terms include data, mapping of object to axis and variables, aestethics of graph (geom) and statistical transformation (stat)

```{r, message=FALSE, warning=FALSE}

require(tidyverse)

#examples 

ggplot() +
layer(
data = diamonds, mapping = aes(x = carat, y = price),
geom = "point", stat = "identity", position = "identity"
) +
scale_y_continuous() +
scale_x_continuous() +
coord_cartesian()


# qplot is a simplified version of ggplot2 using some default inputs

qplot(carat, price, data = diamonds)


```

If a function from a specific package needs to be called, do it in the form ` package::function() `


## mpg dataset examples

```{r, message=FALSE, warning=FALSE}

ggplot2::mpg


```

```{r, eval=FALSE}
?mpg
```

**Displ** variable is engine size
**hwy** is efficiency in galons

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ, y=hwy))

```

First argument (layer) to `ggplot()` is the **Data** argument in the graph grammar. It would be an empty graph without more parameters. 

`geom_point` specifies that aestethics will be points, like shown in the  scatterplot. Each `geom_function` needs a `mapping=aes()` argument to specify variables to axis like x or y. 

R for data science template for graphs:

```{r, eval=FALSE}

ggplot(data = <DATA>) + 
    GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

```

####Exercises

`glimpse()` lets you quickly look at the # of rows and columns in a dataset

```{r, eval=TRUE}

#empty graph
ggplot(data = mpg)

str (mpg) # rows # columns

# mpg$drv variable is type of drive in the cars, front (f), 4 (4wd) or rear (r) 

```

Scatterplot of hwy vs cyl

```{r}


ggplot(data = mpg)+
  geom_point(mapping = aes(hwy, cyl))

```

Scatterplot of class vs drv

```{r}

ggplot(data = mpg)+
  geom_point(mapping = aes(class, drv))

```

Both unuseful due to low number of points and being categorical variables, scatterplot shows relationship between continuous variables. 

## Aesthetics

A visual property of the objects in our graph. 
If we assign a graphical property to class of car in `mpg` graph of `hwy vs. displ` we can see if there are differences according to class (compact, SUV, pickup)

Assigning a unique level of aesthetic to each different value in the variable is known as **scaling**.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```
 Legends do not need to be specified if they can be inferred by `R` directly from the input.
 
Scale of the graph is automatically generated. 
 
If mapped to size, points size will change according to `class` variable

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Mapped to `shapes` and `alpha` (transparency)

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
# maximun 6 shapes, SUVs disappeared

```

Aestethics can be mapped manually to an arbitrary value. This is done outside `aes()` in the following way

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color= "red")

```

#### Exercises

```{r}
# types of variables inside mpg
str(mpg)

# Continuous variable to color produces an scale of the same color varying in intensity 

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=cty))

# Continuous variable to size produces a continually increasing size, proportional to value

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size=cty))

# map same variable to multiple aesthetics

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size=displ))


#stroke aesthetic just changes volume of borders
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, stroke=1))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, stroke=3))

# If you input a conditional statement for an aesthetic, it will assign levels accordingly 
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=displ<5))

```

## Facets

Instead of adding variables to **aesthetics**, you can visualize them as different graphs. **Facets** allows to split plots. 

To use facet in a single variable use `facet_wrap(~ <Variable name>)`.

In the following example is according to `class` variable (needs to be categorical or discrete)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# nrow divides graphs in 2 rows, optional
# ncol can be used to specify # of columns
```

If needs to be a combination of variables use `facet_grid (<Variable 1> ~ <Variable 2>)`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv~ class)
```

- Overall, facets facilitate interpretation as you may judge conditions individually. It also focuses your attention to the item at hand. 
- Especially useful if high number of observations to avoid overlap. 
- A single graph with different aestethics may facilitate graphical comparison

#### Exercises

Facet in a numerical variable
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~cty)

# changes to categorical and creates a graph for each category
```

Replacing one of the values inside `facet_grid (x ~ y)` for ` . ` ignores that axis in the facet (division). See example first for x-axis and then for y-axis. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ drv)

```

## Geometric objects

A **geom** is what a plot uses to represent data: bins, boxes, lines, smotth lines, etc.

All `geom` in ggplot require `mapping` and `aestethics`. 

- All aestethics are not compatible with all **geoms** 

```{r}

ggplot(data = mpg) + 
  geom_point(  mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(  mapping = aes(x = displ, y = hwy))

# A representation with dots or with a smooth fitted line
```

A smooth line can assign a different type of line aestethic to a variable. It cannot fit a shape.  

```{r}
ggplot(data = mpg) + 
  geom_smooth(  mapping = aes(x = displ, y = hwy, linetype= drv))

```

Some `geoms` use a single object to display different rows of data. You can group them by using the aestethic `group`

```{r}
              
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

# group does not add a legend    

ggplot(mpg) +
  geom_smooth(
    mapping = aes(displ, hwy, color = drv))


# You can expect r to recognize data=mpg, x=displ & y=hwy 

# remember to always specify uncommon arguments
```

You can add multiple _geom_ to the same plot. Each one represents a layer. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

```

You can map aestethics directly to the `ggplot()` call in the first layer. Different layers will use  this mapping. 

```{r}

ggplot(data=mpg, mapping = aes(x=displ, y=hwy)) + 
  geom_point() +
  geom_smooth()

```

You can add different aestethics in secondary layers.

```{r}
ggplot(data=mpg, mapping = aes(x=displ, y=hwy)) + 
  geom_point(mapping = aes(color=class)) +
  geom_smooth()

# or use different data

ggplot(data=mpg, mapping = aes(x=displ, y=hwy)) + 
  geom_point(mapping = aes(color=class)) +
  geom_smooth( data = filter(mpg, class=="subcompact"), se=FALSE)

# se defines if standard error shadow is displayed together with fitted line
```

#### Exercises

`show.legend = FALSE` avoids legend .

Recreate different types of graphs in the mpg data

```{r}

#1
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(se=FALSE, color="blue")

#2
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(mapping = aes(line=drv), se=FALSE, color="blue")

#3
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color=drv)) + 
  geom_point() + 
  geom_smooth(se=FALSE)

#4
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color=drv)) + 
  geom_smooth(se=FALSE)


#5
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color=drv)) + 
  geom_smooth(mapping = aes(linetype=drv), se=FALSE)

#6
ggplot(mpg, mapping = aes(displ, hwy)) + 
  geom_point(size=3, color="white") + 
  geom_point(mapping = aes(color=drv))


```


## Statistical transformations

- `Diamonds` data set includes information about 54000 diamonds and their different qualities `(cut, price, carat or weight, clarity and color)` 

```{r}

ggplot(data=diamonds)+
  geom_bar(mapping = aes(x=cut))

#Y axis represents counts or number of diamonds in the cut category, a variable not included in dataset

table(diamonds$cut)
```

- Bars graph need to calculate new values, in this case the ` counts ` of each `cut` type.

- This process is called a **`stat`** or data transformation. You can see the pre-specified stat of a function in the **help section**. `?geom_bar`


- **Stats**  can be used interchangeably with **geom**. They are related by their default association. 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y=..prop.., group=1))

#group in proportions is just a dummy variable so it does not calculate proportions in each strata

ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )

# summarizes the transformation (range and median)

```


#### Exercises

- Same summary graph with default **geom**

```{r}

ggplot(data = diamonds) + 
  geom_pointrange(mapping = aes(x = cut, y = depth) , stat = "summary")

# default summary is mean and sd

ggplot(data = diamonds) + 
  geom_pointrange(mapping = aes(x = cut, y = depth) , stat = "summary",
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median)
```

```{r}
ggplot(data = diamonds) + 
  geom_col(mapping = aes(x = cut, y = depth))

# geom_col used identity of the y-axis variable as stat to construct bars
```

- The *`geom_`* and *`stat_`* associated by default usually share same name  

- If you do not input `group` in geom_bar, it will count each bar and compare it with itself. Each bar will be 100% 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))

```

- With the `fill` aesthetic, the heights of the bars need to be normalized.

- In the example we'll normalize the 

```{r}

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))


ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y= ..count../sum(..count..), fill = color))
```

## Position adjustment


- Bar charts can have different border colors with `color` aestethic or filled with color using `fill`. They can even be combined.  

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

```

- Stacking is automatical. it depens on the position argument.

- Default is `position="identity"` which overlaps in a bar plot. (If color is transparent (alpha), you can see the effect)

- Compare it with `position="fill"` which makes every bar the same highness, facilitating comparisson of percentages along the X-axis

- Dodge option puts overlapping objects inmediately beside the other 

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

**Jitter** can be used in other **geom** to avoid overlapping. It adds a small amount of random noise in the graph so you can see the bulk of dots in a scatterplot.  

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(displ, hwy))

# compare

ggplot(data = mpg) + 
  geom_point(mapping = aes(displ, hwy), position = "jitter")
```

#### Exercises

**_geom_jitter_** variation can be specified with width= or height=

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +   geom_point()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +   geom_point() +
  geom_jitter(width = 0.25)

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +   geom_point() +
  geom_jitter(height = 14)

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +   geom_point() +
  geom_jitter(width = 2, height = 0)

```

- If you use *jitter*, the exact position will slightly change. If you use **_geom_count_** it will adjust the size of the dot according to the count. 
+ It can overlap if counts are too big

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```


- A box plot is used to show distribution across groups. 

+ `Position=dodge2` moves boxplots in a horizontal way only

+ It will separate horizontally any difference between X-axis variable.  

```{r}
?geom_boxplot()

table(mpg$cty)

ggplot(mpg, mapping = aes(hwy, cty))+
  geom_boxplot(aes(group=cut_width(hwy, 5))) 

# can be used with a continuous variable but you have to input a cut to the width like in previous example

ggplot(mpg, mapping = aes(class, hwy, color=drv))+
  geom_boxplot() 

```

## Coordinate systems

Determine the coordinate system (cartesian) where the graph is located. 

- `coord_flip` switches X and Y axis

```{r}

ggplot(mpg, mapping = aes(class, hwy, color=drv))+
  geom_boxplot() +
  coord_flip()
```

- `labs` assign a text label as indicated in the following
+ Assigns label to X and Y initial specification, even if flipped

```{r}
ggplot(mpg, mapping = aes(class, hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    y = "Highway MPG",
    x = "Class",
    title = "Highway MPG by car class",
    subtitle = "1999-2008",
    caption = "Source: http://fueleconomy.gov"
  )

# optional, you can add less 
ggplot(mpg, mapping = aes(class, hwy)) +
  geom_boxplot() +
  labs(
    y = "Highway MPG",
    x = "Class of car",
    title = "Highway MPG by car class"
  )

```

- `?coord_fixed` creates a standar ratio of 1 between X and Y axis
- `geom_abline()` shows a lineal relationship with slope 1 and intercept 0 as defaults

```{r}
?geom_abline()


p <-  ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() 

# with and without fixed ratio of coordinates

p

p + coord_fixed()
```

## Layered grammar of graphs


```{r, eval=F, echo=T}

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

