---
title: "cm007 Exercises: Practice with `dplyr`"
output: 
  html_document:
    keep_md: true
    theme: paper
    toc: true
editor_options: 
  chunk_output_type: console
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tsibble))
```

### Group_by ()


- **`group_by()`** groups a tibble according to the instruction\

    + _Operations you do on a grouped tibble are done independently in each group_
    + Inside `group_by` you can define a new variable and use it to subset

```{r}
gapminder %>% 
  group_by(continent, year)
# Dataframe is not rearranged, grouping appears over output

gapminder %>% 
  group_by(smallLifeExp = lifeExp < 60)
```

### Summarize ()

**`Summarize()`** creates new columns too. However, the results of the calculations are collapsed to _only 1 value_\

- `Summarize` respects previous grouping
        + Results are presented according to the order in which grouping variables where entered\
        
- - `Summarize` also drops any variable not used in calculations

```{r}
gapminder %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

 # compare against different order
gapminder %>% 
  group_by(year, continent) %>%   
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

```

- `dplyr::n()` summarizes the number of rows in each group

```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(n = n())
```

-`dplyr::count()` just counts the number of observations according to a variable

```{r}
gapminder %>% 
  count(country)
```

### Grouped mutate

- `lag()` evaluates the argument inmediately prior in the list according to the grouping variable

```{r}
gap_inc <- gapminder %>% 
  arrange(year) %>% 
  group_by(country) %>%
  mutate(gdpPercap_inc = gdpPercap - lag(gdpPercap))

DT::datatable(gap_inc)
# remember datatable presents data in a nice way
```

- `Mutate` is ideal for vectorized calculations: `mean(), cos(), round()`\

    + Also for window functions `lag(), rank()` that use all data available in a given vector for calculation


`drop_na()` drops any row that has an NA value, in this case all those corresponding to year 1952

```{r}
gap_inc %>% tidyr::drop_na()

```
 
### Using lubridate

Allows easy handling of dates

```{r}
lubridate::mdy("September 24, 2019")
lubridate::mdy("Sep 24 2019")
lubridate::mdy("9-24-19")
lubridate::dym(c("24-2019, September", "25 2019 Sep"))
#Notice that they display the dates all in ymd format, which is best for computing because the dates sort nicely this way.

lubridate::ymd("2019 September 24") %>% 
  class()
# You can tag on hms, too:
lubridate::ymd_hms("2019 September 24, 23:59:59")

```

Other date functions in the 'tsibble' package

```{r}
tsibble::yearmonth("2019 September 24")
tsibble::yearweek("2019 September 24")
```

### Tsibble dataframes

Used for data that contains time series, compare it as tibble and tsibble

- It determines the intervals between values
    + Avoid hidden NA with fill_gaps()
    
```{r}
energy <- read_csv("daily_consumption.csv")

energy

(energy <- as_tsibble(energy, index=date))
# ordered

(energy <- fill_gaps(energy))
```

Gapminder as a tsibble 

```{r}
gapminder %>% 
  as_tsibble(index = year, key = country)
# Key informs grouping
```

Grouping a tsibble needs the index_by instead of group_by

- Can use functions from the lubridate package

```{r}
require(lubridate)

energy %>% 
  tsibble::index_by(day_of_year = yday(date)) %>% 
  dplyr::summarize(mean_intensity = mean(intensity, na.rm = TRUE))
```


## Worksheet

This worksheet contains exercises aimed for practice with `dplyr`. 


1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.

```{r}

gapminder %>% 
  group_by(continent, year) %>% 
  summarize(minlife=min(lifeExp),
          country=country[lifeExp==minlife]) %>%
  arrange(minlife)

```


2. Calculate the growth in population since the first year on record _for each country_ by rearranging the following lines, and filling in the `FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
DT::datatable()
group_by(country) %>% 
```

```{r}
gapminder %>% 
  group_by(country) %>% 
  arrange(year) %>% 
  mutate(rel_growth =  pop - first(pop)) %>%
  DT::datatable()

# first takes the first value in the dataset or according to groups
```


3. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent, sorted by the drop, by rearranging the following lines of code. Ensure there are no `NA`'s. Instead of using `lag()`, use the convenience function provided by the `tsibble` package, `tsibble::difference()`:

```
drop_na() %>% 
ungroup() %>% 
arrange(year) %>% 
filter(inc_life_exp == min(inc_life_exp)) %>% 
gapminder %>% 
mutate(inc_life_exp = FILL_THIS_IN) %>% 
arrange(inc_life_exp) %>% 
group_by(country) %>% 
group_by(continent) %>% 
knitr::kable()
```

```{r}
gapminder %>% 
  group_by(country) %>%
  arrange(year) %>% 
  mutate(inc_life_exp = difference(lifeExp)) %>%
  drop_na() %>% 
  ungroup() %>% 
  group_by(continent) %>% 
  filter(inc_life_exp == min(inc_life_exp)) %>% 
  arrange(inc_life_exp)

```



