---
title: 'cm006: `dplyr` Exercise'
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: inline
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

**Optional, but recommended startup**:

1. Change the file output to both html and md _documents_ (not notebook).

2. `knit` the document. 

3. Stage and commit the rmd, and knitted documents.

# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages. Hint: `suppressPackageStartupMessages()`!
    - This loads `dplyr`, too.
    
##  **_Never use magic numbers to subset as this is not reproducible_**
- Use `%>%` or subset functions     
    
```{r load packages, warning = FALSE, message = FALSE}
# load your packages here:
library("gapminder")
library("tidyverse")
```
    

## `select()` (8 min)

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

```{r}
gapminder %>% select(year,lifeExp, country) %>% head()

```


2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)

# Better way:
select(gapminder, country:lifeExp)
```


3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, -lifeExp)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
# everything() selects all variables, continent before is selected first

select(gapminder, continent, everything()) %>% head()
```

5. Rename `continent` to `cont`.
    + First input is the new variable

```{r}
# select drops all additional variable, rename keeps them
select(gapminder, cont=continent) %>% head()

rename(gapminder, cont=continent) %>% head()
```

## `arrange()` (8 min)

1. Order by year.

```{r}
arrange(gapminder, year)
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, desc(year))
```

3. Order by year, then by life expectancy.

```{r}
arrange(gapminder, year, lifeExp)
```

## Piping, `%>%` (8 min)

Note: think of `%>%` as the word "then"!

Demonstration:

Here I want to combine `select()` Task 1 with `arrange()` Task 3.

This is how I could do it by *nesting* the two function calls:

```{r nesting functions example, eval = F}
# Nesting function calls can be hard to read
arrange(select(gapminder, year, lifeExp, country), year, lifeExp)
```

Now using with pipes:

```{r}
# alter the below to include 2 "pipes"
gapminder %>% select(year, lifeExp, country) %>% 
  arrange(year, lifeExp)
```

# Resume lecture 

Return to guide at section 6.7.


## `filter()` (10 min)

1. Only take data with population greater than 100 million.

```{r}

gapminder %>%
  filter(pop>(100*10^6)) %>% head()
```

2. Your turn: of those rows filtered from step 1., only take data from Asia.

```{r}
gapminder %>%
  filter(pop>(100*10^6)) %>% 
  filter(continent=="Asia")

gapminder %>%
  filter(pop>(100*10^6) & continent=="Asia") %>% head()
```

3. Repeat 2, but take data from countries Brazil, and China. 

    + Operator **`%in%`** is used when evaluating multiple options inside the same variable

```{r}
gapminder %>%
  filter(country== 'Brazil' | country== 'China') %>% 
  head()
```

```{r, eval=FALSE}
gapminder %>% 
  filter(country  %in% c('Brazil', 'China')) %>% 
  head()
```

## `mutate()` (10 min)

The `mutate()` function _creates_ new columns in the tibble by transforming other variables. Like `select()`, `filter()`, and `arrange()`, the `mutate()` function also takes a tibble as its first argument, and returns a tibble. 

The general syntax is:

```
mutate(tibble, NEW_COLUMN_NAME = CALCULATION)
```

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
gapminder %>%
  mutate(GDP=gdpPercap*pop) %>% 
  head()

gapminder %>%
  mutate(GDP=gdpPercap*pop) %>% 
  mutate(
    gdpBill= round(GDP/(1*10^7), 2)) %>% 
  head()
```

Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>%
  transmute(GDP=gdpPercap*pop) %>% 
  transmute(
    gdpBill= round(GDP/(1*10^7), 2)) %>% 
  head()
```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
gapminder %>% 
  mutate(lifeExp=if_else(country=="Canada" & year==1952, 70, lifeExp)) %>%
  filter(country=='Canada') %>% 
  head()

```

Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
gapminder %>% 
  mutate(cc=paste(country, continent, sep = ", ")) %>% head()

```

These functions we've seen are called __vectorized functions__.

## git stuff (Optional)

Knit, commit, push!

# Bonus Exercises

If there's time remaining, we'll practice with these three exercises. I'll give you 1 minute for each, then we'll go over the answer.

1. Take all countries in Europe that have a GDP per capita greater than 10000, and select all variables except `gdpPercap`. (Hint: use `-`)
      + Gapminder data includes countries data

```{r}
gapminder %>% 
  filter(gdpPercap>10000) %>% 
  select(-gdpPercap) %>% 
  head()
```


2. Take the first three columns, and extract the names.

```{r}
gapminder %>% 
  select(1:3) %>% 
  names()
```


3. Of the `iris` data frame, take all columns that start with the word "Petal". 
    - Hint: take a look at the "Select helpers" documentation by running the following code: `?tidyselect::select_helpers`.
    
```{r}
iris <- as_tibble(iris)
# dplyr works better with tibbles

iris %>% 
  select(starts_with("Petal"))


```

    
4. Convert the population to a number in billions.

    + North american billion equals to 1*10^7
```{r}
gapminder %>% 
  mutate(pop.billions=pop/(1*10^7)) %>% 
  head()
```


5. Filter the rows of the iris dataset for Sepal.Length >= 4.6 and Petal.Width >= 0.5.

```{r}
iris %>% 
  filter(Sepal.Length >= 4.6 & Petal.Width >= 0.5)
```


Exercises 3. and 5. are from [r-exercises](https://www.r-exercises.com/2017/10/19/dplyr-basic-functions-exercises/).