---
title: "cm007 Exercises: Practice with `dplyr`"
output: 
  html_document:
    keep_md: true
    theme: paper
    toc: true
editor_options: 
  chunk_output_type: console
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tsibble))
```

### Group_by ()


- **`group_by()`** groups a tibble according to the instruction\

    + _Operations you do on a grouped tibble are done independently in each group_
    + Inside `group_by` you can define a new variable and use it to subset

```{r}
gapminder %>% 
  group_by(continent, year)
# Dataframe is not rearranged, grouping appears over output

gapminder %>% 
  group_by(smallLifeExp = lifeExp < 60)
```

### Summarize ()

**`Summarize()`** creates new columns too. However, the results of the calculations are collapsed to _only 1 value_\

- `Summarize` respects previous grouping
        + Results are presented according to the order in which grouping variables where entered\
        
- - `Summarize` also drops any variable not used in calculations

```{r}
gapminder %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

 # compare against different order
gapminder %>% 
  group_by(year, continent) %>%   
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))

```

- `dplyr::n()` summarizes the number of rows in each group

```{r}
gapminder %>% 
  group_by(country) %>% 
  summarize(n = n())
```

-`dplyr::count()` just counts the number of observations according to a variable

```{r}
gapminder %>% 
  count(country)
```

### Grouped mutate

- `lag()` evaluates the argument inmediately prior in the list according to the grouping variable

```{r}
gap_inc <- gapminder %>% 
  arrange(year) %>% 
  group_by(country) %>%
  mutate(gdpPercap_inc = gdpPercap - lag(gdpPercap))

DT::datatable(gap_inc)
# remember datatable presents data in a nice way
```

- `Mutate` is ideal for vectorized calculations: `mean(), cos(), round()`\

    + Also for window functions `lag(), rank()` that use all data available in a given vector for calculation

## Worksheet

This worksheet contains exercises aimed for practice with `dplyr`. 


1. (a) What's the minimum life expectancy for each continent and each year? (b) Add the corresponding country to the tibble, too. (c) Arrange by min life expectancy.


2. Calculate the growth in population since the first year on record _for each country_ by rearranging the following lines, and filling in the `FILL_THIS_IN`. Here's another convenience function for you: `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
DT::datatable()
group_by(country) %>% 
```

```{r}

```


3. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent, sorted by the drop, by rearranging the following lines of code. Ensure there are no `NA`'s. Instead of using `lag()`, use the convenience function provided by the `tsibble` package, `tsibble::difference()`:

```
drop_na() %>% 
ungroup() %>% 
arrange(year) %>% 
filter(inc_life_exp == min(inc_life_exp)) %>% 
gapminder %>% 
mutate(inc_life_exp = FILL_THIS_IN) %>% 
arrange(inc_life_exp) %>% 
group_by(country) %>% 
group_by(continent) %>% 
knitr::kable()
```

```{r}

```



