---
title: "cm012 Exercises: Factors"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    theme: paper
editor_options: 
  chunk_output_type: console
---

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r, message=FALSE, echo=FALSE}
packages <- c(
  "gapminder", 
  "tidyverse", 
  "forcats",
  "ggplot2")
lapply(packages, require, character.only = TRUE)
```

## Motivating the need for factors in R

Factors are how we store categorical variables in `R`.\
\
The values a factor can take are called **levels** and are showed as strings ("Colombia", "Venezuela") altough each *level* is really assigned a number: 1,2,3,etc.

- Check your variables when something weard happens, use `class()`
- Base read.csv(), read.table() and data.frame() change *strings* to *factors* by default

### Forcats package

Special package to handle factor related operations

```{r}
str(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)
```

```{r}
fct_count(gapminder$continent)
```

Even if you drop multiple values in a factor variable, the levels do not change

```{r}
nlevels(gapminder$country)

h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>% 
  filter(country %in% h_countries)
nlevels(h_gap$country)
```

```{r}
# use of fct_drop () removes unused levels
h_gap$country %>% 
  fct_drop() %>% 
  levels()
```

Factors are usually ordered alphabetically, you can change it accordingo to some principle

-fct_infreq() orders by decreasing frequency
-fct_rev() reverts the order

```{r}
gapminder$continent %>% 
  fct_infreq() %>%
  levels()

# alphabetically but backwards
gapminder$continent %>%
  fct_rev() %>% 
  levels()

# increasing frequency
gapminder$continent %>% 
  fct_infreq() %>%
  fct_rev() %>% 
  levels()
```

- fct_reorder() lets you order factors according to other variable.
    + Useful for better plots
    + The default summarizing function parsed to the index variable is *median ()*
    + Other function can be specified
    
```{r}
fct_reorder(gapminder$country, gapminder$lifeExp)  %>%
  levels() %>%
  head()

## order accoring to minimum life exp instead of median
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% 
  levels() %>%
  head()

## backwards median
fct_reorder(gapminder$country, gapminder$lifeExp, .desc = TRUE) %>% 
  levels() %>% 
  head()
```
    
- fct_Recode() changes the *string* assigned to levels in a factor

```{r}
i_gap <- gapminder %>% 
  filter(country %in% c("United States", "Sweden", "Australia")) %>% 
  droplevels()
i_gap$country %>%
  fct_recode("USA" = "United States", "Oz" = "Australia") %>% levels()

i_gap <- gapminder %>% 
  filter (country %in% c("Australia", "Korea, Dem. Rep.", "Korea, Rep.")) %>% 
  droplevels()
i_gap$country %>% 
  fct_recode("Oz" = "Australia", "North Korea" ="Korea, Dem. Rep.", "South Korea"= "Korea, Rep.") %>% 
  levels()

```

### Factors for plotting

Use gapminder data from 1997 and create a new factor varaible based on life expectancy

```{r}
gapminder %>% 
  filter(year==1997) %>% 
  mutate(life_level = factor(case_when(
    lifeExp < 48~ 'low',
    lifeExp < 59 ~ 'moderate',
    lifeExp < 70 ~ 'high',
    TRUE ~ 'very high'))) %>%
  ggplot() +
  geom_boxplot(aes(x = life_level, y = gdpPercap)) +
  labs(y = "GDP per capita, $", x= "Life expectancy level, years") +
  theme_bw() 
```

In called to factor, we can set the levels we want and avoid droping levels in the X-axis.

```{r}
gapminder %>% 
  filter(year == 1997) %>% 
  mutate(life_level = factor(case_when(lifeExp < 23 ~ 'very low',
                                lifeExp < 48~ 'low',
                                lifeExp < 59 ~ 'moderate',
                                lifeExp < 70 ~ 'high',
                                TRUE ~ 'very high'),
                levels = c("very low" , "low","moderate", "high","very high"))) %>%  
  ggplot() + geom_boxplot(aes(x = life_level, y = gdpPercap)) +
  labs(y = "GDP per capita, $", x= "Life expectancy level, years") +
  theme_bw() +
  scale_x_discrete(drop = FALSE)
```

