---
title: "Tidy data"
output: 
  html_document:
    keep_md: true
    toc: true
    theme: paper
---

# Tidy Data

In tidy data:
-Each variable forms a column.
-Each observation forms a row.
-Each cell contains a value
```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

```{r}
browseVignettes("tidyr")
```

## Pivoting
### Pivot_longer

`pivot_longer` makes datasets longer by increasing the number of rows and decreasing the number of columns.

- `cols` column names to be modified
- `names_to` old column names are going to a new column. What is the name(s) of the(se) new columns?
- `values_to` similar but with data contained in value cells, what is the name of the(se) new columns
```{r, message=FALSE}
require(tidyverse)

relig_income
relig_income %>% 
  pivot_longer(cols = -religion, 
               names_to = "income", 
               values_to = "count")
```

- `names_prefix` specifies a character string to remove from start of each variable name
- `names_ptypes = list()` changes variable types
```{r}
billboard
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )

billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    names_prefix = "wk",
    names_ptypes = list(week = integer()),
    values_to = "rank",
    values_drop_na = TRUE,
  )
```

- `names_pattern` can be used when a column contains multiple values. 
    + `new_?` ignores the string
```{r}
who
who %>% 
  pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count"
)
# ?tidyr::extract
```

When multiple observations are included per row
- Use **`.value`**: specify whic part of the column name will remain as a column (variable) 
    + In this case, `names_pattern` or `names_sep` defines which corresponds to `.value` and `set`
```{r}
anscombe
anscombe %>% 
  pivot_longer(everything(), 
    names_to = c(".value", "set"), 
    names_pattern = "(.)(.)"
  ) %>% 
  arrange(set)
```

### Pivot_wider

Makes a dataset wider by increasing the number of columns and decreasing the number of rows. 

- `id_cols=` what you want to keep as it is
- `names_from=`Here you specify from what column do the new column names come , can be multiple columns
- `values_from:` the values under the new columns are coming from an old column. Which column is this?
```{r}
fish_encounters
fish_encounters %>% 
  pivot_wider( id_cols = fish,
    names_from = station,
    values_from = seen)
```

- You can specify if new variables will include a transformation (data has 9 replication)
```{r}
warpbreaks <- warpbreaks %>% 
  as_tibble() %>% 
  select(wool, tension, breaks) %>%
  print()
warpbreaks %>% 
  pivot_wider(
    names_from = wool, 
    values_from = breaks,
    values_fn = list(breaks = mean)
  )
```

## Making tibbles

Cannot recycle and argument unless it is of length 1
```{r}
tibble(x = 1:6,
       y = 1:2)
```

Using `tibble::tribble()`
```{r}
tribble(
  ~Day, ~Breakfast,
  1, "Apple",
  2, "Yogurt",
  3, "Yogurt"
)
```

Can also nest dataframes inside columns
```{r}
(list_col <- tibble(n = 1:2,
                    y = list(iris, mtcars)))

(iris_nest <- iris %>% 
  group_by(Species) %>% 
  nest())
```

`expand_grid()` expands all possible combinations
```{r}
expand_grid(x = 1:2, y = 1:2, z = 1:2)
```

#### Hidden `NA`
```{r}
(df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
))
```

If we look at all the possible combinations, some `NA` appear
```{r}
df %>% 
  expand(year, qtr) 

df %>% 
  complete(year, qtr) #makes explicit NA's
```


