---
title: "cm011-File input/Output (I/O)"
output: 
  html_document:
    toc: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  warning = FALSE, error = TRUE, message = FALSE)
```

## General concepts

- _Surprise me!_ is the best idea, a lot of cleanining and mingling is expected every time you import a dataset.
- **_Try to make the most out of your importing function_**
- Export to known and simple text files
- _Modify data with commands_, is more functional in the long run
- Create separated scripts with intermediate output files.
    + Helps you start from intermediate steps

## Readr package

Readr does not convert string to factors automatically.\
\
Includes multiple `parse` functions, each works by column of imported data (heuristic guess).\
\
Base function is `read_delim()` with specific functions for text `read_tsv` and comma separated files `read_csv()`
- Using these functions saves lots of time in cleaning data, **lots**

```{r, eval=FALSE}
browseVignettes("readr")
# Introduction to readr
```

```{r}
require(tidyverse)
require(gapminder)
require(here)
```

### Writing data to disk

```{r}
gap_asia_2007 <- gapminder %>% 
  filter(year == 2007, continent == "Asia")
gap_asia_2007
```

Write as `.csv`
```{r}
write_csv(gap_asia_2007,"exported_file.csv") #default location is wd of .rmd file

```

## **here** package

When inside an `R.proj` file, the working directory is always at the level of the file.\
\
If you are working in a `.rmd` file. The working directory is where the file is located.
- Create a separate `data` sub-directory\
\
`here::here()` shows your `wd` and can help you specify locations in relationship to it using strings

```{r}
here::here()
# directory with my R.proj file
here::here("data")
```

- Specify sub-folders by adding strings 
```{r}
here::here("data", "cm011_data", "05Nov2019")
# not actually created unless you put some data in
write_csv(gap_asia_2007,here::here("Weeks_5-6", "data.cm011", "exported_file.csv"))
here()
```

- Using it with read functions
```{r}
read_csv(here::here("Weeks_5-6","data.cm011","exported_file.csv"))
```

### Reading data from the web

Specify the path as the web URL
```{r}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
read_csv(url)
```

