---
title: "cm010 Exercises: Tibble Joins"
output: 
  html_document:
    toc: true
    theme: paper
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  warning = FALSE, error = TRUE, message = FALSE)
```

## Join tibbles

Databases are usually presented divided in multiple tables related through some key variable. This is named relational databases. 
```{r}
browseVignettes("dplyr") # two-table verbs
require(tidyverse)
```

If you use `bind`, the rows or columns will be coerced in order, independently of match. **Beware**

### Mutating joins

Add new columns to the tibble. 
- Match databases according to their *variables* and add new variables (columns).
- They return `NA` to fill remaining rows, depending on type of merge.
-`by=` can specify variables to use for match, e.g. `by="mangaka"` or `by=c('mangaka'='author')`

#### Types of mutating joins

1. `inner_join (x,y)` includes only observations with match in both
2. `left_join (x,y)` returns all observations in **x**
3. `right_join (x,y)` is similar but keeps all in **y**
4. `full_join (x,y)` keeps all rows of both
```{r}
df1 <- tibble(x = c(1, 2), y = 2:1)
df2 <- tibble(x = c(1, 3), a = 10, b = "a")

df1 %>% inner_join(df2) 

df1 %>% left_join(df2) 

df1 %>% right_join(df2) 

df1 %>% full_join(df2) 
```

### Filtering joins

They join `tibbles or tables` based on variables but affect mainly the observations (rows usually) instead of.
- Filtering joins never duplicate data. 
- `semi_join(x,y)` returns all the rows in x that **match** y
- `anti_join(x,y)` return all the rows in x **without a match** in y
```{r}
df1 <- tibble(x = c(1, 1, 3, 4), 
              y = 1:4)
df2 <- tibble(x = c(1, 1, 2), 
              z = c("a", "b", "a"))

df1 %>% nrow() # Four rows to start with

df1 %>% inner_join(df2, by = "x") # And we get four rows after the join

df1 %>% semi_join(df2, by = "x") ## But only two rows actually match
```

_**Joins** are conservative, if a **factor variable** doesn't have the same levels and in the same order_ 