---
title: "cm010 Exercises: Tibble Joins"
output: 
  html_document:
    theme: readable
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set (warning = FALSE, error = TRUE)
```


## Requirements

You will need Joey's `singer` R package for this exercise. And to install that, you'll need to install `devtools`. Running this code in your console should do the trick:

```
install.packages("devtools")
devtools::install_github("JoeyBernhardt/singer")
```

If you can't install the `singer` package, we've put the singer data on the `STAT545-UBC/Classroom` repo, and you can load those in instead by running the following lines of code:

```
songs <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/singer/songs.csv")
locations <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/singer/loc.csv")
```

Load required packages:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(singer)
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning = FALSE, fig.align = "center")
```

## Exercise 1: `singer`

The package `singer` comes with two smallish data frames about songs. Let's take a look at them (after minor modifications by renaming and shuffling):

```{r}
(time <- as_tibble(songs) %>% 
   rename(song = title))
```

```{r}
(album <- as_tibble(locations) %>% 
   select(title, everything()) %>% 
   rename(album = release,
          song  = title))
```

1. We really care about the songs in `time`. But, which of those songs do we know its corresponding album?

```{r}
time %>% 
  FILL_THIS_IN(album, by = FILL_THIS_IN)

time %>% 
  semi_join(album, by = c("song","artist_name"))
# keeps only the rows in time that have matches in album
```

2. Go ahead and add the corresponding albums to the `time` tibble, being sure to preserve rows even if album info is not readily available.

- `left_join` keeps all rows in time
    + `by=NULL` is the default and joins by all the posible matches in Col_names
```{r}
time %>% 
  FILL_THIS_IN(album, by = FILL_THIS_IN)

time %>% 
  left_join(album)
```

3. Which songs do we have "year", but not album info?

- `anti_join` keeps all data in `time` not matched in `album`
```{r}
time %>% 
  FILL_THIS_IN(album, by = "song")

time %>% 
  anti_join (album, by = "song")
```

4. Which artists are in `time`, but not in `album`?

- Same but using `artist_name`
```{r}
time %>% 
  anti_join(album, by = "artist_name")
```


5. You've come across these two tibbles, and just wish all the info was available in one tibble. What would you do?

```{r}
time %>% 
  full_join (album, by = c("song","artist_name"))
```


## Exercise 2: LOTR

Load in the three Lord of the Rings tibbles that we saw last time:

```{r}
fell <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
ttow <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
retk <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

1. Combine these into a single tibble.

```{r}
FILL_THIS_IN(fell, FILL_THIS_IN)

bind_rows(fell, ttow, retk)
```

2. Which races are present in "The Fellowship of the Ring" (`fell`), but not in any of the other ones?

```{r}
fell %>% 
  anti_join (ttow, by = "Race") %>% 
  FILL_THIS_IN(FILL_THIS_IN, by = "Race")

fell %>% 
  anti_join (ttow, by = "Race") %>% 
  anti_join (retk, by = "Race")
# No races
```

## Exercise 3: Set Operations

Let's use three set functions: `intersect`, `union` and `setdiff`.
We'll work with two toy tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
```

```{r}
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

1. Rows that appear in both `y` and `z`

```{r}
FILL_THIS_IN(y, z)
intersect(y, z)
inner_join(y,z)
?intersect
```

2. You collected the data in `y` on Day 1, and `z` in Day 2. Make a data set to reflect that.

```{r}
union(
  mutate(y, day = "Day 1"),
  mutate(z, day = "Day 2")
)
# or bind_rows
```

3. The rows contained in `z` are bad! Remove those rows from `y`.

```{r}
setdiff(y,z)
anti_join(y,z)
```
